Design and Simulation of an gene expression experiment
======================================================

# Questions
- is gene expression different in differentiated cells?
- monocytes differentiates to macrophages
- drug concentrations induce gene expression?

# Design

## 5 cases
 case | differentiated  | drug concentration  
 ------------- | ------------- | -------------  
 1 | - | 0  
 2  | + | 0  
 3 | + | 2.5  
 4 | + | 5  
 5 | + | 7  


\[R=R_{min}+(R_{max}-R_{min})*P_{logis} \]
\[R_{max}=F_{fold}\times R_{min}\]
\[R=R_{min}(1+(F_{fold}-1))*P_{logis}(hill*log(conc/EC_{50})) \]

## Seed
R generates pseudo rondom numbers. You can get identical sets of random numbers with set.seed(1234). With varying seed you will get different results for each simulaton step.

## Confidence intervals
Confidence intervals should include the start parameter set!

```{r start}
project<-"design"
myhome<-"~/Arbeitsfläche"
setwd(paste(myhome,"simStudy",project,sep="/"))
getwd()
```

5 genes and duplicate measurement of each sample ID

```{r packages,message=FALSE}
library("ProjectTemplate")
load.project()
opts_chunk$set(cache=FALSE,dev = c('png','pdf'),echo=FALSE, message=FALSE, fig.width=11,fig.height=8,warning=FALSE,results='hide')
library(MASS)
library(nlme)
source("src/myFun.R")
source("src/helpers.R")
```

```{r parameter}
selGene<-c("HPRT",paste("ABCD",1:3,sep=""))
mv<-c(HPRT=1000,ABCD1=1000,ABCD2=10,ABCD3=1000)
conc<-c(0,0.1,2.5,5,7)
diff<-c(FALSE,rep(TRUE,length(conc)-1))
diff<-c("D","O",rep("D",length(conc)-2))
hill=20
fold<-c(HPRT=1.,ABCD1=1.3,ABCD2=5,ABCD3=1/3)
ec50<-c(HPRT=10,ABCD1=5,ABCD2=5,ABCD3=5)
eps<-.1^2
omega<-.2^2
Sigma<-diag(rep(omega,length(mv)))

```


# Conc response function

- between sample variability \( \omega \) = `r signif(sqrt(omega),3)`
- within sample or residual variability \( \sigma \) = `r signif(sqrt(eps),3)`

parameter | HPRT | ABCD1 | ABCD2 | ABCD3
 -------- | ----- |----- | ----- | -----
$R_{min}$ | 1000  | 1000 | 10 | 1000
$F_{fold}$ | 1  | 1 | 5 | .7
$EC_{50}$ | 1  | 1 | 5 | 5


```{r makeData}
nRep<-3
nObs<-2
dd<-expand.grid(Case=seq(along=conc), Rep=1:nRep,Obs=1:nObs)
dd

dd<-transform(dd,ID=factor(paste("S", Case,Rep,sep=""),ordered=T),Conc=conc[Case],FCase=LETTERS[Case])
xtabs(~ID,dd)
dd<-dd[order(dd$ID,dd$Rep,dd$Obs),]
dd
yy<-function(x,fa,ec50) (1+(fa-1)*plogis(hill*log(x/ec50)))
sgene<-selGene[3]
xyplot(yy(0:9,fold[sgene],ec50[sgene])~0:9,scale=list(x=xyLogscale()),type="b")

#yy(0:9,fold[sgene],ec50[sgene])
```



```{r makeRandomVector}
#set.seed(1234);
head(dd)
xtabs(~ID,dd)
rand<-with(dd,mvrnorm(length(levels(ID)),rep(0,length(mv)),Sigma=Sigma)[ID,])
dimnames(rand)<-list(dd$ID,names(mv))
mvm<-matrix(mv,ncol=length(mv),nrow=nrow(dd),byrow=T,dimnames=dimnames(rand) )
ddr<-cbind(dd,mvm*exp(rand))
ddrm<-myMelt(ddr,selGene,varn="Gene",valn="RNA")

```


```{r teil2,results='markup'}
names(ddrm)[names(ddrm)%in%c("variable","value")]<-c("Gene","RNA")
ddrm<-transform(ddrm, RNAobs=RNA*yy(Conc,fold[Gene],ec50[Gene]))
ddrm$RNAobs<-ddrm$RNAobs*exp(rnorm(nrow(ddrm),0,eps))         
head(ddrm)
```

```{r gd}
gd<-groupedData(RNAobs~Conc|Rep,ddrm)
sapply(gd,class)
plot(gd,scale=list(y=xyLogscale()),aspect="fill",outer=~Gene,key=T)
plot(gd,outer=~Gene,aspect="fill",scale=list(y=c(xyLogscale(c(1,5E4)),list(relation="free"))),key=T)

```



```{r gdF}
gd<-groupedData(RNAobs~Gene|ID,ddrm)
pp<-
  plot(gd,displayLevel=1,collapseLevel=1,,scale=list(x=xyLogscale(c(1,1E4))),aspect="fill",inner=~Gene,mv=mv)
pp$panel<-function (x, y, subscripts, groups,mv) 
{
    dot.line <- trellis.par.get("dot.line")
    panel.abline(h = y, v=log10(mv) ,lwd = dot.line$lwd, lty = dot.line$lty, 
        col = dot.line$col)
    panel.superpose(x, y, subscripts, groups)
}
pp

```

# lme

```{r fit1}
sapply(gd,class)
head(gd)
gd$lobs<-log10(gd$RNAobs)
levels(gd$Gene)
levels(gd$FCase)
fms1<-lme(lobs~Gene*FCase,gd,random=~1|ID)
fms2<-update(fms1,random=~(Gene-1)|ID)
fms3<-update(fms1,random=list(ID=pdDiag(~(Gene-1))))
anova(fms3,fms2,fms1)
anova(fms3,fms2)
fmsfin<-fms3
anova(fmsfin)
summary(fmsfin)$tT
gdInt<-intervals(fmsfin,levels=.68)

```


```{r compare,results='markup'}
#gdInt$fix
10^gdInt$fix#[1:4,]
mv
gdInt$reStruct
gdInt$sigma
c(omega=sqrt(omega),eps=sqrt(eps))
```



# barplot


```{r barplot}
int0<-gdInt$fix
int<-int0[grep("^Gene",rownames(int0)),]
int
rownames(int)<-gsub("FCase","",gsub("Gene","",rownames(int)))
sel<-grep(":",rownames(int))
int<-int[sel,]

tmp<-unPaste(rownames(int),sep=":")
tmp
inta<-cbind(data.frame(int),Gene=tmp[[1]],FCase=tmp[[2]])
intal<-daply(inta,.(Gene),function(x){
  y<-as.matrix(x[,1:3]);
  dimnames(y)[[1]]=x$FCase
  y})
#daply(inta,.(Gene),function(x) {y<-as.matrix(x[,1:3]);y})

intal
#barPlotsInt(intal,T)
x<-intal
ym <- x[,,2];yup <-  x[,,3];ylow <-  x[,,1]
LOG=F
if(!LOG) {ym <-10^ym;yup <- 10^yup;ylow <- 10^ylow}
#LAS=2;lay=NULL;ylim=NULL;col=0
#barx <- barplot(ym,beside=T,col=col,las=LAS,ylim=ylim,ylab=ylab,legend.text=T,args.legend = list(x = leg,ncol=2,bty = "n"),layout=lay)
ylim<-c(0,max(yup)*1.1)
col=rainbow(length(levels(gd$Gene))-1)
barx <- barplot(ym,beside=T,col=col,las=1,legend.text=T,args.legend = list(x = "topleft",ncol=1,bty = "n"),ylim=ylim)
abline(h=1)
error.bar(barx,ym,yup-ym,ym-ylow)

```

```{r transpose}
ym<-t(ym);yup <- t(yup);ylow <- t(ylow)
col=rainbow(length(levels(dd$FCase))-1)
barx <- barplot(ym,beside=T,col=col,las=1,legend.text=T,args.legend = list(x = "topleft",ncol=1,bty = "n"),ylim=ylim)
abline(h=1)
error.bar(barx,ym,yup-ym,ym-ylow)
```


# Results

- no influence of cell differentiation is assumed
- lme results represents input parameter
- experimental design sufficient to show strong induction of gene expression
- repeating simulations with different assumption may result in the sensitivity of the design.


